Best practices for forecasting changes in biodiversity: an example using breeding birds

### 2017-09-05 manuscript/draft.Rmd
Discussion edits

The accuracy for forecasts is expected to decline as forecasts are made further
into the future, so it is important to evaluate forecasts at different lead
times [Box 1; @petchey2015]. For deviance (the error metric that evaluates both
models' point estimates and their uncertainty simultaneously), there is some
indication that the JSDM and the nonlinear richness model converge towards the
time-series and baseline models as the lead time of the forecast increases. This
potential convergence is biologically plausible because at short lead times,
environmental changes are relatively small and changes in the biota may lag
behind changes in the environment. However, as the lead time of the forecast
increases the amount of environmental change should increase and the systems
will have more time to respond, potentially resulting in shifts towards improved
relative performance of models incorporating environmental and ecological
information [@blonder2017]. This is potentially important for interpreting our
results because we only evaluated annual forecasts up to a decade into the
future. In contrast, many ecological forecasts are temporally aggregated to 5-30
year timesteps and have lead times of up to a century. These long-term forecasts
are difficult to evaluate given the small number of century scale datasets, but
are important for understanding changes in biodiversity at some of the lead
times relevant for conservation and management. Two studies have assessed models
of species richness at longer lead times [@algar2009; @distler2015], but the
results were not compared to baseline or time-series models (in part due to data
limitations) making them difficult to compare to our results. Given the
potential changes with lead time of the models and processes are most relevant
to forecasting, it is important that more long lead time evaluation studies are
conducted and that the inferences drawn from short-term and long-term
forecasting studies be compared. For true forecasts, short term forecasts will
allow for relatively rapid validation of models by researchers without needing
to wait decades for data to accumulate [@tredennick2016; @dietze2016]. However,
since drivers of species richness differ at different temporal scales
[@rosenzweig1995; @white2004; @white2007], short-term forecasts should be viewed
as informing, not replacing, long-term forecasts [@tredennick2016].

While it is possible that models that include species' relationships to their
environments or direct environmental constraints on richness will provide better
fits at longer lead times, it is also possible that they will continue to
produce forecasts that are worse than baselines assuming that the systems are
static.  This would be expected to occur if richness in these systems is not
changing over the relevant multi-decadal time scales, which would make simpler
models with no directional change more appropriate. Recent suggestions that
local scale richness in some systems is not changing directionally at
multi-decadal scales supports this possibility [@brown2001; @ernest2001;
@vellend2013; @dornelas2014]. A lack of change in richness may be expected even
in the presence of substantial changes in environmental conditions and species
composition at a site due to replacement of species from the regional pool
[@brown2001; @ernest2001]. On average, the Breeding Bird Survey sites used in
this study show little change in richness [site-level SD of `r numbers$resid_sd`
species, after controlling for differences among observers; see also
@lasorte2005]. The absence of rapid change in this dataset is beneficial for the
absolute accuracy of forecasts across different sites: when a past year's
richness is already known, it is easy to estimate future richness.  The
site-level stability of the BBS data also explains why SDMs and macroecological
models perform relatively well at predicting future richness, despite failing to
capture changes in richness over time. However, this stability also makes it
difficult to improve forecasts relative to simple baselines, since those
baselines are already close to representing what is actually occurring in the
system. These results suggest that single-site models should be actively
considered for forecasts of richness and other stable aspects of
biodiversity. Our results also suggest that future efforts to understand and
forecast biodiversity should incorporate species composition, since lower-level
processes are expected to be more dynamic [@ernest2001; @dornelas2014] and
contain more useful information [@harris2015].

Future biodiversity forecasting efforts also need to address the uncertainty
introduced by the error in forecasting the environmental conditions that are
used as predictor variables. In this, and other hindcasting studies, the
environmental conditions for the "future" are known because the data has already
been observed. However, in real forecasts the environmental conditions
themselves have to be predicted, and environmental forecasts will also have
uncertainty and bias. Ultimately, ecological forecasts that use environmental
data will therefore be more uncertain than our current hindcasting efforts. It
is important to correctly incorporate this uncertainty in the predictor
variables into forecasting models [@clark2001; @dietze2017]. Limitations in
forecasting future environmental conditions, particularly at small scales, will
present continued challenges for models incorporating environmental variables,
and this may result in a continued advantage for simple single-site approaches.

In addition to comparing and improving the process models used for forecasting
it is important to consider the observation models.  When working with any
ecological dataset, there are imperfections in the sampling process that have
the potential to influence results. With large scale surveys and citizen science
datasets, such as the Breeding Bird Survey, these issues are potentially
magnified by the large number of different observers and by major differences in
the habitats and species being surveyed [@sauer1994]. Accounting for differences
in observers reduced the average error in our point estimates and also improved
the coverage of the confidence intervals. In addition, controlling for observer
effects resulted in changes in which models performed best, most notably
reducing the relative performance of the naive model's point estimates. This
suggests that, prior to accounting for observer effects, the naive model
performed well because it was capable of accommodating rapid shifts in estimated
richness introduced by changes in the observer. These kinds of rapid changes
were difficult for the other single-site models to accommodate, and so the
average performance of the ARIMA and average models improved once this source of
observation error was addressed. This demonstrates that modeling observation
error can be important for properly estimating and reducing uncertainty in
forecasts and can also lead to changes in the best methods for forecasting [Box
1]. Another key aspect of an ideal observation model is imperfect detection. In
this study, we did not address differences in detection probability across
species and sites [@boulinier1998] since there is no clear way to address this
issue using NA Breeding Bird Survey data without making strong assumptions about
the data [i.e., assuming there is no biological variation in stops along a
route; @white2010], but this would be a valuable addition to future forecasting
models.

The science of forecasting biodiversity remains in its infancy and it is
important to consider weaknesses in current forecasting methods in that
context. In the beginning weather forecasts were also worse than simple
baselines, but these forecasts have continually improved throughout the history
of the field [@mcgill2012; @silver2012; @bauer2015]. One practice that lead to
improvements in weather forecasts was that large numbers of forecasts were made
publicly, allowing different approaches to be regularly assessed and refined
[@mcgill2012; @silver2012]. To facilitate this kind of improvement, it is
important for ecologists to start regularly making and evaluating real
ecological forecasts, even if they perform poorly, and to make these forecasts
openly available for assessment [@mcgill2012; @dietze2016]. These forecasts
should include both short-term predictions, which can be assessed quickly, and
mid- to long-term forecasts, which can help ecologists to assess long time-scale
processes and determine how far into the future we can successfully forecast
[@dietze2016; @tredennick2016]. Forecasts out to 2050 based on the models in
this paper have been openly archived [@zenodo], so that we and others can assess
how well they perform. We plan to evaluate these forecasts and report the
results as each new year of BBS data becomes available, and make iterative
improvements to the forecasting models in response to these assessments.

>>>>

We have only evaluated annual forecasts up to a decade into the future, but
forecasts are often made with a lead time of 50 years or more. These long-term
forecasts are difficult to evaluate given the small number of century-scale
datasets, but are important for understanding changes in biodiversity at some of
the lead times relevant for conservation and management. Two studies have
assessed models of species richness at longer lead times [@algar2009;
@distler2015], but the results were not compared to baseline or time-series
models (in part due to data limitations) making them difficult to compare to our
results directly. Given the potential changes with lead time of the models and
processes are most relevant to forecasting, it is important that more forecasts
with longer lead times are evaluated and that the inferences drawn from
short-term and long-term forecasting studies be compared. By evaluating
shorter-term forecasts, researchers can avoid waiting so many decades to
evaluate their methods [@petchey2015; @tredennick2016; @dietze2016]. However,
since drivers of species richness differ at different temporal scales
[@rosenzweig1995; @white2004; @white2007], short-term evaluations should be
viewed as informing, not replacing, long-term evaluations [@tredennick2016]. In
particular, we might expect environmental and ecological information to become
more important at longer time scales, and thus for the performance of simple
baseline forecasts to degrade faster than forecasts from SDMs and other similar
models [@blonder2017].  On the other hand, there is evidence that species
respond differently to spatial versus temporal variation in the environment
[@oedekoven2017], implying that most SDMs are not using environmental
information well enough to make accurate forecasts.

While it is possible that models that include species' relationships to their
environments or direct environmental constraints on richness will provide better
fits at longer lead times, it is also possible that they will continue to
produce forecasts that are worse than baselines that assume the systems are
static.  This would be expected to occur if richness in these systems is not
changing over the relevant multi-decadal time scales, which would make simpler
models with no directional change more appropriate. Recent suggestions that
local scale richness in some systems is not changing directionally at
multi-decadal scales supports this possibility [@brown2001; @ernest2001;
@vellend2013; @dornelas2014]. A lack of change in richness may be expected even
in the presence of substantial changes in environmental conditions and species
composition at a site due to replacement of species from the regional pool
[@brown2001; @ernest2001]. On average, the Breeding Bird Survey sites used in
this study show little change in richness [site-level SD of `r numbers$resid_sd`
species, after controlling for differences among observers; see also
@lasorte2005]. The absence of rapid change in this dataset is beneficial for the
absolute accuracy of forecasts across different sites: when a past year's
richness is already known, it is easy to estimate future richness.  The
site-level stability of the BBS data also explains why SDMs and macroecological
models perform relatively well at predicting future richness, despite failing to
capture changes in richness over time. However, this stability also makes it
difficult to improve forecasts relative to simple baselines, since those
baselines are already close to representing what is actually occurring in the
system. These results suggest that single-site models should be actively
considered for forecasts of richness and other stable aspects of biodiversity.
Our results also suggest that future efforts to understand and forecast
biodiversity should incorporate species composition, since lower-level processes
are expected to be more dynamic [@ernest2001; @dornelas2014] and contain more
useful information [@harris2015].

Future biodiversity forecasting efforts also need to address the uncertainty
introduced by the error in forecasting the environmental conditions that are
used as predictor variables. In this, and other hindcasting studies, the
environmental conditions for the "future" are known because the data has already
been observed. However, in real forecasts the environmental conditions
themselves have to be predicted, and environmental forecasts will also have
uncertainty and bias. Ultimately, ecological forecasts that use environmental
data will therefore be more uncertain than our current hindcasting efforts, and
it is important to correctly incorporate this uncertainty into our models
[@clark2001; @dietze2017]. Limitations in forecasting future environmental
conditions---particularly at small scales---will present continued challenges
for models incorporating environmental variables, and this may result in a
continued advantage for simple single-site approaches.

In addition to comparing and improving the process models used for forecasting
it is important to consider the observation models.  When working with any
ecological dataset, there are imperfections in the sampling process that have
the potential to influence results. With large scale surveys and citizen science
datasets, such as the Breeding Bird Survey, these issues are potentially
magnified by the large number of different observers and by major differences in
the habitats and species being surveyed [@sauer1994]. Accounting for differences
in observers reduced the average error in our point estimates and also improved
the coverage of the confidence intervals. In addition, controlling for observer
effects resulted in changes in which models performed best, most notably
improving most models' point estimates relative to the naive baseline [Box 1].
This suggests that, prior to accounting for observer effects, the naive model
performed well largely because it was capable of accommodating rapid shifts in
estimated richness introduced by changes in the observer. These kinds of rapid
changes were difficult for the other single-site models to accommodate. Another
key aspect of an ideal observation model is imperfect detection. In this study,
we did not address differences in detection probability across species and sites
[@boulinier1998] since there is no clear way to address this issue using North
American Breeding Bird Survey data without making strong assumptions about the
data [i.e., assuming there is no biological variation in stops along a route;
@white2010], but this would be a valuable addition to future forecasting models.

The science of forecasting biodiversity remains in its infancy and it is
important to consider weaknesses in current forecasting methods in that context.
In the beginning, weather forecasts were also worse than simple baselines, but
these forecasts have continually improved throughout the history of the field
[@mcgill2012; @silver2012; @bauer2015]. One practice that lead to improvements
in weather forecasts was that large numbers of forecasts were made publicly,
allowing different approaches to be regularly assessed and refined [@mcgill2012;
@silver2012]. To facilitate this kind of improvement, it is important for
ecologists to start regularly making and evaluating real ecological forecasts,
even if they perform poorly, and to make these forecasts openly available for
assessment [@mcgill2012; @dietze2016]. These forecasts should include both
short-term predictions, which can be assessed quickly, and mid- to long-term
forecasts, which can help ecologists to assess long time-scale processes and
determine how far into the future we can successfully forecast [@dietze2016;
@tredennick2016]. We have openly archived forecasts from all six models through
the year 2050 [@zenodo], so that we and others can assess how well they perform.
We plan to evaluate these forecasts and report the results as each new year of
BBS data becomes available, and make iterative improvements to the forecasting
models in response to these assessments.
